Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-24 index.qmd
section:background; process:write

  Dichotomous Comparative Judgment [DCJ, @Pollitt_2012a, @Pollitt_2012b] requires judges to evaluate the relative manifestation of traits between pairs of stimuli, resulting in a dichotomous outcome indicating which stimulus exhibits the trait more strongly. Research has demonstrated DCJ's effectiveness and reliability in various domains [@Pollitt_2012b; @Bartholomew_et_al_2018; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Boonen_et_al_2020]. Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that generate the DCJ data. Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data. This study aims to fill this gap by applying the framework of causal analysis and Directed Acyclic Graphs [DAG; @Pearl_2009]. Using this framework, the study will construct a scientific model to elucidate the causal assumptions and mechanisms inherent the system. This model will enable researchers to draw inferences about causal relationships from DCJ data. Subsequently, the study will translate this model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference. The outcomes of this study will inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.

@@

<!-- 5. final objective and relevance -->
Ultimately, this research aims to extend the law of comparative judgment initially proposed by @Thurstone_1927 and provide a sound probabilistic base for the statistical analysis of DCJ data. Consequently, this research holds significance for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.

@@

<!-- 1. What is causal inference? -->
In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical representation for analyzing causes and counterfactuals [@Pearl_2009].

<!-- 2. what are counterfactuals? -->
According to @Pearl_et_al_2018, counterfactuals occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association, and form the foundation of causal inference. Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in the absence of those causes [@Pearl_et_al_2018]. Counterfactuals achieves this by translating causal statements to statements about "what would have happened under different circumstances."

@@

<!-- 4. But approach are we going to use? -->
However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. These graphical representations provide a transparent depiction of a system’s complexity, revealing its underlying assumptions and the plausible causal mechanisms responsible for generating the system’s outcome. Ultimately, DAGs play the important role of effectively conveying the assumed causal structure of the system, and also determine wich variables should be controlled for to be able to estimate causal effects. Consequently, the structural approach will be the primary method used to address one of the study’s research goals.

@@

<!-- 1. What are DAGs? and how are they represented? -->
As previously mentioned, the structural approach employs Directed Acyclic Graphs (DAGs) for the graphical representation of systems. A *graph* consists of nodes connected by edges [@Gross_et_al_2018]. In a *directed graph*, edges extend from a *parent* node to a *child* node, with arrows indicating the direction of causal influence. In directed *acyclic* graphs, the direction of causal influences does not loop back on themselves, meaning the graph contains no cycles [@Neal_2020, @McElreath_2020].

@@

<!-- 8. What are the DAG's building blocks?, and what do they imply? (SCMs) -->

>>>>

  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait. DCJ has proven effective and reliable across various fields @Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020]. However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data are generated. This study addresses these issues by integrating DCJ into the framework of causal inference. Specifically, utilizing a structural approach to causal inference, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system. It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal connections within the system. This research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data. The findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.

@@

<!-- 5. final objective and relevance -->
Ultimately, this research builds upon Thurstone’s law of comparative judgment [-@Thurstone_1927] and provide a robust probabilistic foundation for the statistical analysis of DCJ data. Consequently, the findings of this study offer valuable insights for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.

@@

<!-- 1. What is causal inference? -->
In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical notation for analyzing causes and counterfactuals [@Pearl_2009].

<!-- 2. what are counterfactuals? -->
According to @Pearl_et_al_2018, counterfactuals occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association, and form the foundation of causal inference. Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."

@@

<!-- 4. But which approach are we going to use? -->
However, the structural approach offers a key advantage over the potential outcomes approach: it enables the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. In this context, DAGs fulfill the role of a heuristic, effectively conveying the assumed causal structure of a system. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAG is accurate [@McElreath_2020].

<!-- A Directed Acyclic Graph (DAG) allows for the transparent depiction of a system’s complexity, revealing its underlying assumptions and the plausible mechanisms generating the system’s outcome, assuming the DAG is true.  -->

@@

<!-- 1. What are DAGs? and how are they represented? -->
Graph theory is the branch of mathematics focused on the study of graphs. Graphs are mathematical structures used to model pairwise relations between objects [@Gross_et_al_2018]. While graph theory covers a wide array of topics, the field of causal inference, particularly its structural approach, has incorporated some of its concepts to represent causes and counterfactuals formally and transparently. A causal graph, or Directed Acyclic Graph (DAG), as its name suggest, is a directed graph without cycles. A *graph* is a collection of nodes connected by edges. In a *directed graph*, edges extend from a node to another node, with arrows indicating the direction of causal influence. In a directed *acyclic* graph, the direction of causal influences does not loop back on itself, ensuring that the graph contains no cycles [@Neal_2020, @McElreath_2020].

@@

<!-- 8. What are the DAG's building blocks?, and what do they imply? (SCMs) -->
<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->
